Experience Report: Haskell in Computational Biology

### 2012-06-28 mrfy_experience_report.tex
wordsmithing section 3.2

\newenvironment{smallverbatim}{\par\small\verbatimsmalltrue\verbatim}{\endverbatim}
\newcommand\smallverbatiminput[1]{%
  \everypar=\expandafter{\the\everypar
       \verbatimsmallfalse
       \belowdisplayskip=\abovedisplayskip
       \belowdisplayshortskip=\abovedisplayshortskip
       \topsep=\standardvspace}%
  \topsep=0.78\topsep
  \belowdisplayskip=0.78\belowdisplayskip
  \belowdisplayshortskip=0.78\belowdisplayshortskip
  \verbatimsmalltrue\verbatiminput{#1}}
\newcommand\smallverbatiminputx[1]{\vspace*{-8pt}{\small\verbatiminput{#1}}} % WTF?? {\LaTeX} hell
\newcommand\smallfuzzverbatiminput[2]{{\par\unskip\small\hfuzz=#1 \verbatiminput{#2}}}

@@

Our implementations of hidden Markov models and Viterbi's algorithm
duplicate existing functionality; these same algorithms were
implemented in C++ for HMMER and SMURF.
This reimplementation is a cost of using Haskell, and one that
other researchers may have to pay as well, as fewer computational biology
libraries exist for Haskell. For example, the National Center for
Biotechnology Information publishes its own library of algorithms
for computational biology.
However, the authors of SMURF and HMMER implemented these algorithms
anew, even though existing implementations existed in C++.
In~MRFy, we use these algorithms as building blocks to implement
new functionality.

A hidden Markov model uses local information about
amino-acid sequences.
But when a real protein folds in three dimensions, 
groups of amino acids 
that are far away in the one-dimensional sequence can be
adjacent in three-dimensional space.
We~are studying adjacent groups called \emph{beta strands}, which
can become hydrogen-bonded to each other,
and are effectively ``stuck together.''

A~mutation in a paired beta strand may break the pairing, in which
 case the mutated protein will not fold properly and is unlikely to
 survive natural selection.
But a mutation in a paired beta strand \emph{can} survive if it is
 accompanied by a \emph{compensatory} mutation in the strand with which it is
 paired. 
The probability of finding amino acid~$x_i$ in column~$j$ of an alignment may
 therefore depend on the amino acid in the paired column
 $\pairedwith j$.
This nonlocal interaction changes
Viterbi's equations:
each term $V_{j}^{\prime M}(i)$ becomes
$$W_{j}^{M}(i) = V_{j}^{\prime M}(i) - \log P(x_{i} \mid x_{\pairedwith j}).$$
The distance between columns~$j$ and $\pairedwith j$ can be as small
 as a few columns or as large as a few hundreds of columns.

The approach currently taken in computational biology
is~to model each pair of beta strands not with two hidden Markov
models, 
but with a pair of sequences of match states.
Each of the subsequences \emph{between} paired beta strands is then
modeled with an ordinary hidden Markov model.
The~combined model, an~example of which is shown in \figref{mrf}, is called a
\textit{Markov random field}. 
Because of nonlocal interactions between paired beta strands, 
the likelihood-maximization problem for a Markov random field does not
 exhibit optimal substructure, and dynamic
 programming cannot solve it quickly \cite{Menke:2010ti,Daniels:2012}.
MRFy explores new methods of \emph{approximating}
maximum likelihood.

MRFy treats the beta 
strands, which are built into the model,
 as ``beads'' which can be slid along the query sequence.
A~set of positions of beta strands is called
 a \emph{placement}.
A~placement's probability is computed
based on observed frequencies of paired amino acids in 
hydrogen-bonded beta strands~\cite{Cowen:2002p588}.
Given a placement, the maximum likelihood of any subsequence that appears
\emph{between} two beta strands can be computed exactly and
efficiently using Viterbi's algorithm.
(The result of the exact computation is a probability that is
\emph{conditioned} on the placement.)

>>>>

\newenvironment{smallverbatim}{\par\small\verbatimsmalltrue\verbatim}{\endverbatim}
\newcommand\smallverbatiminput[1]{%
  \everypar=\expandafter{\the\everypar
       \verbatimsmallfalse
%       \belowdisplayskip=\abovedisplayskip
%       \belowdisplayshortskip=\abovedisplayshortskip
       \topsep=\standardvspace}%
  \topsep=0.78\topsep
%  \belowdisplayskip=0.78\belowdisplayskip
%  \belowdisplayshortskip=0.78\belowdisplayshortskip
  \verbatimsmalltrue\verbatiminput{#1}}
\newcommand\smallverbatiminputx[1]{\vspace*{-8pt}{\small\verbatiminput{#1}}} % WTF?? {\LaTeX} hell
\newcommand\smallfuzzverbatiminput[2]{{\par\unskip\small\hfuzz=#1 \verbatiminput{#2}}}

@@

Our Haskell code for hidden Markov models and Viterbi's algorithm
solves the same problems as existing C++~code.
Other researchers using Haskell may also have to reimplement code,
but researchers may also reimplement code out of choice, not
necessity.
For~example, both SMURF and HMMER contain new implementations of these
same algorithms.
In~MRFy, we are using the Haskell implementations to help solve new
problems.

When a real protein folds in three dimensions, 
amino acids 
that are far away in the one-dimensional sequence can be
adjacent in three-dimensional space.
We~are studying groups of such acids called \emph{beta
strands}.\remark{Fix this; we're not really studying beta strands}
Beta strands
can become hydrogen-bonded to each other,
making them ``stuck together.''
MRFy detects homologous proteins that include hydrogen-bonded beta
strands; using prior methods, many instances of this problem are
intractable.

To~account for beta strands, computational biologists must use both a
richer model of protein structure and a revised version of Viterbi's
equations.
The~richer model recognizes that only certain pairs of amino acids are
able to bond to each other.
%%  The presence of beta strands changes both the natu
%%   the probability that a given
%%  query sequence fit
%%  A~mutation in a paired beta strand may break the pairing, in which
%%   case the mutated protein will not fold properly and is unlikely to
%%   survive natural selection.
%%  But a mutation in a paired beta strand \emph{can} survive if it is
%%   accompanied by a \emph{compensatory} mutation in the strand with which it is
%%   paired. 
When column~$j$ of an alignment is part of a beta strand and is paired
with another column  $\pairedwith j$,
the probability of finding amino acid~$x_i$ in column~$j$ 
depends on~$x_{\pairedwith j}$.
The~relevant term in
Viterbi's equations 
$V_{j}^{\prime M}(i)$ becomes
$$W_{j}^{M}(i) = V_{j}^{\prime M}(i) - \log P(x_{i} \mid x_{\pairedwith j}).$$
Columns~$j$ and $\pairedwith j$ can be as close together as a few
columns or as far apart as a few hundreds of columns.

The presence of beta strands also changes the model.
Within a beta strand, amino acids are not inserted or deleted;
a~bonded pair of beta strands is modeled by
 a pair of sequences of match states.
Between the beta strands, the model is structured as before.
The~combined model, an~example of which is shown in \figref{mrf}, is called a
\textit{Markov random field}. 
Becayuse $W_j^M(i)$~depends not only on $x_i$ and its local neighbors but
also on the nonlocal $x_{\pairedwith j}$, 
dynamic programming cannot compute the maximum likelihood quickly
\cite{Menke:2010ti,Daniels:2012}.
MRFy explores new methods of \emph{approximating}
maximum likelihood.

MRFy treats the beta 
strands in the model
 as ``beads'' which can be slid along the query sequence.
A~positioning of the beta strands is called
 a \emph{placement}.
A~placement's probability is computed
based on observed frequencies of paired amino acids in 
hydrogen-bonded beta strands~\cite{Cowen:2002p588}.
Given a placement, the maximum likelihood of the remaining parts of
the query sequence, which appear between and around beta strands,
can be computed quickly and exactly 
using Viterbi's algorithm.
The exact result is a probability that is
\emph{conditioned} on the placement.
